The demographic costs of sexually antagonistic selection in partially selfing populations

### 2021-06-03 doc/SA-Hermaphrodites-wDemography.tex
Minor edits
Edited first paragraph of Results to hopefully hold demographers by the hand a bit more when looking at the funnel plots, and as a consequence moved some info around in the results.

Shortened methods by one paragraph.

Added few sentences to discussion.

For a given parameter set, we numerically evaluate the largest eigenvalue of the Jacobian evaluated at both boundaries, and the population growth rate of the equilibrium obtained after perturbation of the boundary, $\lambda$. To determine the viability threshold (where the population growth rate equals one) and the boundary stability thresholds (where the largest eigenvalue of the Jacobian crosses one), we start at the outer boundaries of selection parameter space, and titrate inward. The proportions of total $s_f \times s_m$ parameter space where SA polymorphism and/or extinction occurred were calculated by numerically integrating the appropriate regions defined by the invasion and extinction thresholds.

For analyses of the demographic effects of inbreeding depression, we make two additional simplifying assumptions: First, we assume that if inbreeding depression is caused primarily by recessive deleterious mutations (as suggested by empirical data), it should covary negatively with the population selfing rate due to purging, provided the population selfing rate has been relatively constant in recent evolutionary time (\citealt{Charlesworth2009}; though we note that other processes could give rise to this pattern, e.g., \citealt{CrnokrakBarrett2002, Charlesworth2009,HedrickGarcia-Dorado2016}). Following \citet{Olito2019}, we incorporate such negative covariance by constraining the inbreeding depression terms in the model ($\delta$ and $\delta_i$, where $i \in \{j,a,\gamma\}$) to follow a simple declining function of the selfing rate: $\delta = \delta^{\ast} (1 - b (1 - L))$, where $\delta^{\ast}$ is the hypothetical severity of inbreeding depression for a completely outcrossing population, $b$ is a shape parameter determining how far $\delta$ will decline under complete selfing (when $C = 1$), and $L$ describes the expected deleterious mutation load as a function of the selfing rate $C$. The function $L$ includes an additional shape parameter, $a$, which determines the curvature of the overall function for $\delta$ (see Appendix E in \citealt{Olito2019} for additional details). We set $\delta^{\ast} = 0.8$, $b = 0.5$, and $a = 0.2$ for all analyses, values chosen to be consistent with empirical estimates of inbreeding depression (e.g., fig.~2 in \citealt{HusbandSchemske1996}). Second, to keep our analyses tractable we explore the effects of individual inbreeding depression terms in isolation. That is, we assume that only one of the $\delta$ and $\delta_i$ terms (where $i \in \{j,a,\gamma\}$) can be non-zero at a time.

@@

We begin with an illustration of demographically viable polymorphic parameter space in the absence of inbreeding depression in Figure \ref{fig:extThresholds} (i.e., $\delta = \delta_i = 0$). Invasion conditions for SA alleles in the evolutionary demographic model closely match the predictions from population genetic models \citep{Kidwell1977, JordanConnallon2014, Olito2017}. In particular, the demographic model recovers the classic "funnel-shaped" region of polymorphic $s_f \times s_m$ parameter space. The effects of the population selfing rate ($C$) and dominance ($h$) on SA polymorphism are also similar: self-fertilization ({\itshape i}) reduces the parameter space in which male-benefit alleles can invade, thereby increasing opportunity for spread of female-benefit alleles (e.g., contrast figure 1A with figure 1C); and ({\itshape ii}) dominance reversals (where deleterious SA fitness effects are partially recessive in each sex; $h = 1/4$) are much more permissive of SA polymorphism (e.g., contrast figures 1A-C with 1D-F) \citep{JordanConnallon2014, Olito2017}.

However, a key prediction from the evolutionary demographic model is that large fractions of SA polymorphic parameter space can be demographically inviable (fig.~\ref{fig:extThresholds}). The location of the extinction threshold, where the population intrinsic growth rate $\lambda = 1$, is primarily determined by the fertility parameter ($f$) but is also influenced by the population selfing rate ($C$) and dominance of the SA alleles ($h$). As $f$ increases, the population can survive in the face of stronger fertility selection through both male and female sex-functions before extinction occurs (the extinction threshold shifts to higher values of $s_f$ and $s_m$).

The demographic consequences of SA selection are driven by the fitness effects of male-beneficial/female-deleterious alleles because they directly influence individuals' female fecundity, and therefore population growth rates. For the fertility parameter values we explored, all extinctions occured in regions where either the female-deleterious allele segregates at intermediate frequency (i.e., where SA polymorphism is maintained), or goes to fixation (fig.~\ref{fig:extThresholds}). The SA polymorphisms that remained viable at lower fertilitiy values corresponded to regions where the female-deleterious allele is predicted to segregate at low frequencies, or in regions with weak selection through both sex functions, an asymmetry that reflects the assumption of female demographic dominance.

@@

The first and central finding of our study is that when intrinsic population growth rates approach one, the deleterious effects of segregating male-beneficial SA alleles on female fecundity can result in extinction over much of the parameter space where SA polymorphism is maintained. Since growth rates far exceeding one suggest rapid exponential growth, they are generally rare (with the notable exception of recently introduced, invasive populations), suggesting that our model predictions may be highly relevant for SA polymorphism in many real-world populations. Moreover, these findings complement recent theoretical and empirical studies indicating that SA selection is likely to be both condition dependent, and stronger in locally-adapted populations near the center of a species' range, where population growth rates are expected to be high \citep{BergerEtAl2014, Connallon2015}.

We also find that demographically viable parameter space is often biased towards alleles with stronger selection through the female than male sex function. That is, given natural variation in population growth rates, the most demographically viable (and therefore observable in natural populations) outcomes of SA selection are either fixation of a female-beneficial allele, or polymorphisms involving low-frequency female-deleterious alleles. This key prediction is supported by a series of experimental results in seed beetles ({\itshape Callosobruchus maculatus}) which show that male-beneficial SA genotypes are less likely to contribute positively to population growth rates, and are more susceptible to extinction under environmental stress or inbreeding \citep{BergerEtAl2014, berger2016intralocus, grieshop2017male}. \textcolor{red}{Jessica's comment Re: sex-limited exp. evol.}. An interesting corrollary of our findings is that, since strong SA fitness effects often lead to extinction, the observable SA genetic variation in natural populations may often be under weak selection, and therefore strongly susceptible to genetic drift whether or not selection favours the fixation of one allele or balancing selection \citep{ConnallonClark2012}.

>>>>

% For a given parameter set, we numerically evaluate the largest eigenvalue of the Jacobian evaluated at both boundaries, and the population growth rate of the equilibrium obtained after perturbation of the boundary, $\lambda$. To determine the viability threshold (where the population growth rate equals one) and the boundary stability thresholds (where the largest eigenvalue of the Jacobian crosses one), we start at the outer boundaries of selection parameter space, and titrate inward. The proportions of total $s_f \times s_m$ parameter space where SA polymorphism and/or extinction occurred were calculated by numerically integrating the appropriate regions defined by the invasion and extinction thresholds.

%%% LdV: We say more or less the same thing two paragraphs earlier, so I think we don't need to explain this here anymroe?

For analyses of the demographic effects of inbreeding depression, we make two additional simplifying assumptions: First, to keep our analyses tractable we explore the effects of individual inbreeding depression terms in isolation. That is, we assume that only one of the $\delta$ and $\delta_i$ terms (where $i \in \{j,a,\gamma\}$) can be non-zero at a time. Second, we assume that if inbreeding depression is caused primarily by recessive deleterious mutations (as suggested by empirical data), it should covary negatively with the population selfing rate due to purging, provided the population selfing rate has been relatively constant in recent evolutionary time (\citealt{Charlesworth2009}; though we note that other processes could give rise to this pattern, e.g., \citealt{CrnokrakBarrett2002, Charlesworth2009,HedrickGarcia-Dorado2016}). Following \citet{Olito2019}, we incorporate such negative covariance by constraining the inbreeding depression terms in the model ($\delta$ and $\delta_i$, where $i \in \{j,a,\gamma\}$) to follow a simple declining function of the selfing rate: $\delta = \delta^{\ast} (1 - b (1 - L))$, where $\delta^{\ast}$ is the hypothetical severity of inbreeding depression for a completely outcrossing population, $b$ is a shape parameter determining how far $\delta$ will decline under complete selfing (when $C = 1$), and $L$ describes the expected deleterious mutation load as a function of the selfing rate $C$. The function $L$ includes an additional shape parameter, $a$, which determines the curvature of the overall function for $\delta$ (see Appendix E in \citealt{Olito2019} for additional details). We set $\delta^{\ast} = 0.8$, $b = 0.5$, and $a = 0.2$ for all analyses, values chosen to be consistent with empirical estimates of inbreeding depression (e.g., fig.~2 in \citealt{HusbandSchemske1996}).

@@

We begin with an illustration of demographically viable polymorphic parameter space in the absence of inbreeding depression in Figure \ref{fig:extThresholds} (i.e., $\delta = \delta_i = 0$). The red areas in Figure \ref{fig:extThresholds} indicate areas where the population goes extinct (population growth rate is less than one) for different values of the fertility parameter.  Lower female fertility values correspond to a larger demographically inviable area. The funnel between the thick black lines indicates the area where the sexually antagostic alleles coexist. Above the funnel, the female-benefit and male-detriment allele fixates; below the funnel, the male-benefit and female-detriment allele fixates. Much of the region where the male-benefit and female-detriment allele fixates is demographically inviable due to the demographic consequences of reduced female fitness in this region.  The SA polymorphisms that remain viable at lower fertility values correspond to regions where the female-deleterious allele is predicted to segregate at low frequencies, or in regions with weak selection through both sex functions, an asymmetry that reflects the assumption of female demographic dominance.

Invasion conditions for SA alleles in the evolutionary demographic model closely match the predictions from population genetic models \citep{Kidwell1977, JordanConnallon2014, Olito2017}. In particular, the demographic model recovers the classic "funnel-shaped" region of polymorphic $s_f \times s_m$ parameter space. The effects of the population selfing rate ($C$) and dominance ($h$) on SA polymorphism are also similar: self-fertilization ({\itshape i}) reduces the parameter space in which male-benefit alleles can invade, thereby increasing opportunity for spread of female-benefit alleles (e.g., contrast figure 1A with figure 1C); and ({\itshape ii}) dominance reversals (where deleterious SA fitness effects are partially recessive in each sex; $h = 1/4$) are much more permissive of SA polymorphism (e.g., contrast figures 1A-C with 1D-F) \citep{JordanConnallon2014, Olito2017}.

However, a key prediction from the evolutionary demographic model is that large fractions of SA polymorphic parameter space can be demographically inviable (fig.~\ref{fig:extThresholds}). The location of the extinction threshold, where the population intrinsic growth rate $\lambda = 1$, is primarily determined by the fertility parameter ($f$) but is also influenced by the population selfing rate ($C$) and dominance of the SA alleles ($h$).

% I moved this information up. 
%The demographic consequences of SA selection are driven by the fitness effects of male-beneficial/female-deleterious alleles because they directly influence individuals' female fecundity, and therefore population growth rates. For the fertility parameter values we explored, all extinctions occured in regions where either the female-deleterious allele segregates at intermediate frequency (i.e., where SA polymorphism is maintained), or goes to fixation (fig.~\ref{fig:extThresholds}). The SA polymorphisms that remained viable at lower fertilitiy values corresponded to regions where the female-deleterious allele is predicted to segregate at low frequencies, or in regions with weak selection through both sex functions, an asymmetry that reflects the assumption of female demographic dominance.

@@

The first and central finding of our study is that when intrinsic population growth rates approach one, the deleterious effects of segregating male-beneficial SA alleles on female fecundity can result in extinction over much of the parameter space where SA polymorphism is maintained. Since growth rates far exceeding one suggest rapid exponential growth, they are generally rare (with the notable exception of recently introduced, invasive populations), suggesting that our model predictions may be highly relevant for SA polymorphism in many real-world populations. In addition, we find that much of the parameter space where a male-beneficial allele fixates is demographically inviable for populations whose growth rates are close to one prior to invasion. This demographic consequence of masculinization is rarely considered \citep{hitchcock2020gene}.  Moreover, these findings complement recent theoretical and empirical studies indicating that SA selection is likely to be both condition dependent, and stronger in locally-adapted populations near the center of a species' range, where population growth rates are expected to be high \citep{BergerEtAl2014, Connallon2015}.

We also find that demographically viable parameter space is often biased towards alleles with stronger selection through the female than male sex function. That is, given natural variation in population growth rates, the most demographically viable (and therefore observable in natural populations) outcomes of SA selection are either fixation of a female-beneficial allele, or polymorphisms involving low-frequency female-deleterious alleles. This key prediction is supported by a series of experimental results in seed beetles ({\itshape Callosobruchus maculatus}) which show that male-beneficial SA genotypes are less likely to contribute positively to population growth rates, and are more susceptible to extinction under environmental stress or inbreeding \citep{BergerEtAl2014, berger2016intralocus, grieshop2017male}. An interesting corrollary of our findings is that, since strong SA fitness effects often lead to extinction, the observable SA genetic variation in natural populations may often be under weak selection, and therefore strongly susceptible to genetic drift whether or not selection favours the fixation of one allele or balancing selection \citep{ConnallonClark2012}.

%Comment Jess:  Many sex-limited experimental evolution studies have found a stronger response to male-limited selection compared to female-limited selection. I think people often think that this is because sexual selection is expected to be stronger in males. But it seems like these results support other explanations, e.g. that fixation of male-benefit alleles is only possible in a lab setting where population growth rates are controlled, or maybe that female-benefit alleles should fix more easily and therefore be less likely to segregate in the population.
